File-based localization of numerical perturbations in data analysis pipelines

### 2020-01-06 paper.tex

Operating systems and other computational factors are known to have an
effect on the results produced by neuroimaging pipelines, presumably due to
the creation, propagation and amplification of numerical errors along data
analysis pipelines. However, the precise causes of such instabilities and
the path along which they propagate in the pipelines are unclear, while
this process could play a major role in the current reproducibility crisis.
We present a technique to identify the computational processes in a
pipeline that create numerical differences when executed in different
computational conditions. Our technique leverages system-call interception to reconstruct
processing and file graphs without pipeline instrumentation. Applying our tool to
the structural pre-processing pipelines released by the Human Connectome
Project, we find that linear and non-linear registration are the cause of
most numerical instabilities in such pipelines, which confirms previous
findings.

@@

In neuroimaging, our primary application field, data analyses often consist
of hundreds of computational processes, sometimes coming from multiple
toolboxes, that are aggregated to perform a specific function. For
instance, the fMRIprep pipeline~\cite{esteban2019fmriprep} assembles software blocks
from FSL, AFNI, Freesurfer and ANTs to provide a state-of-the art
functional MRI processing tool with minimal user input. Another example are the pipelines of the Human
Connectome Project~\cite{glasser2013} that combine tools from FSL and Freesurfer to pre-process
structural, functional and diffusion data from their
open dataset. In both cases, the pipelines leverage toolboxes that are
widely trusted in the community. At the same time, substantial results
variations have been observed in these toolboxes resulting from minor data
or infrastructure perturbations~\cite{Glatard2015, Gronenschild2012, Lewis2017-ll}, which suggests that further investigation of their
numerical conditioning is required. For such complex pipelines, a
lightweight solution has to be found to perform such evaluations with
limited code instrumentation.

@@

Figure~\ref{fig:overview} shows an overview of our method. For each tested
condition, Condition 1 and Condition 2, the pipeline is containerized with
Docker, and its interface is described with
Boutiques~\cite{glatard2017boutiques}. \reprozip~\cite{chirigati2016reprozip} is
added to Condition 1 for provenance tracking. Our tool starts by
identifying temporary files and ``multi-write" files, i.e., files written
by more than one process (Figure~\ref{fig:overview-capturing}). To do that,
the tool first executes the pipeline in Condition 1, to obtain a process
graph (Figure~\ref{fig:overview-capturing} - (1)) and a list of result files.
The list of processes that create intermediary or multi-write
files is then generated (Figure~\ref{fig:overview-capturing} - (2)). 
Using this list, the tool modifies Docker image in Condition 1 and
exectutes the pipeline again through it, which turns to produce results
including temporary and overwritten files
(Figure~\ref{fig:overview-capturing} - (3,4)). In the second step, the
processes are labeled from the differences found in their outputs in the
two tested conditions (Figure~\ref{fig:overview-labelling}). To this order,
the tool builds docker image for Condition 2 by modifying pipeline
processes (Figure~\ref{fig:overview-labelling} - (1)) that recognised from 
the process graph. Then, the modified docker
image is executed in Condition 2 and each process is labeled right after
the running by comparing the results obtained in Condition 1
and output files of the process in Condition 2
(Figure~\ref{fig:overview-labelling} - (2)). We then label processes in 2
categories depending whether they
create differences (represented in red) or not (green). 
The tool replaces output file with the same file obtained in Condition 1 
if process creates differences (Figure~\ref{fig:overview-labelling} - (3)).
Process labelling
is done through hooks which are defined during the docker image modifying
procedure, and incremental updates in Condition 2, until the end of the
pipeline execution.

\begin{figure}
  \centering
  \begin{subfigure}{\columnwidth}
    \centering
    \includegraphics[width=1\columnwidth]{images/fig1-a}
    \caption{Identification of temporary and multi-write files.
    (\href{https://docs.google.com/drawings/d/1rsbFwPjNPNMRhXUl4RSK3G0OaiJC7K-buiHN0hPD8eQ/edit?usp=sharing}{Source})}
    \label{fig:overview-capturing}
  \end{subfigure}
   \begin{subfigure}{\columnwidth}
    \centering
     \includegraphics[width=1\columnwidth]{images/fig1-b}
     \caption{Labelling process graphs including intermediary files.
     (\href{https://docs.google.com/drawings/d/1yblYuKWAD18aJe5JxBu2h1EysyoQA3YrsQj-1T3s0l8/edit?usp=sharing}{Source})}
     \label{fig:overview-labelling}
   \end{subfigure}
   \caption{An overview of Spot tool in two steps. First capturing intermediary files (a); then
            labelling process graphs (b).}
   \label{fig:overview}
  \end{figure}

@@

Our tool reconstructs a \emph{process tree} starting from the first process
created by the pipeline and traversing processes through \texttt{clone()}
and \texttt{fork()} dependencies. Our tool also creates a process \emph{graph} 
(Figure~\ref{fig:overview-capturing} - (1)) from this tree
by adding edges corresponding to file dependencies between processes. A
file dependency is defined between processes A and B if a file written by A
is read by B. Figure~\ref{fig:simple_script} shows an example of a process
tree and graph constructed from the example pipeline in
Listing~\ref{listing:sample-script}.
\begin{listing}
\begin{minted}[frame=single,
  framesep=3mm,
  linenos=false,
  xleftmargin=0pt,
  tabsize=4]{bash}
#!/bin/bash

@@

# Run FSL bet, put result ${bet_output}
bet "${input_image}" "${bet_output}" > bet_temp.out
echo "Voxels / Volume in brain mask:"
# Run FSL stats on ${bet_output}
fslstats "${bet_output}" -V
# Run FSL maths on ${bet_output}
fslmaths "${bet_output}" "${bet_output_binarized}"
# Remove temporary file
\rm bet_temp.out
\end{minted}
  \caption{Example pipeline}
  \label{listing:sample-script}
\end{listing}

@@

@@

We label in two categories the processes of the graph of each subject. First,
processes that read files that do not have differences and write files that
do not have differences are labeled \emph{transparent}
(Figure~\ref{fig:green}). Second, processes that read files that do not
have differences but write files that have differences \emph{create}
differences (Figure~\ref{fig:red}). In addition, processes that read files
that have differences and write files that also have differences are called
\emph{undetermined} (Figure~\ref{fig:grey}). Processes are also labeled
undetermined when they read or write temporary files, or produce multi-write
files. To resolve undetermined processes, hooks must be added to the pipeline to 
execute these processes on files without differences, and to capture
intermediary file states.

@@

Files written by multiple processes also lead 
to undetermined labels. For a file F 
written by all processes in \textbf{P} = \{$P_{1}$, \ldots $P_{n}$\}, we 
(1) check that processes in \textbf{P} do not write concurrently to F, 
(2) we establish an order on \textbf{P} based on the creation timestamp 
of the processes, (3) we replace every process $P_{i}$ in \textbf{P} by 
a wrapper that first calls $P_{i}$ and then saves F to a read-only 
directory. Thus, multiple versions of F are saved and used in the 
analysis.

Furthermore, cycles may be present in the process graph in case a file 
was written by more than one process. We remove such cycles by removing 
file edges between processes A and B when A's process creation 
timestamp is posterior to B's or when A=B. Indeed, such edges cannot 
happen in practice unless A and B were running concurrently, which we 
assume is not the case (the tool checks for that).

As an example of this capturing process, in the sample pipeline in 
Listing~\ref{listing:sample-script}, process \texttt{2\#bet2} writes a temporary file and 
also file \emph{T1w\_brain\_bin.nii.gz} is first written by process \texttt{2\#bet2} and 
then overwritten by process \texttt{3\#fslmaths}. The list of these processes is captured in 
Listing~\ref{listing:json_process}.
Using this list, we modify relevant processes in Condition 1 and then re-execute pipeline 
to save them in a backup folder.
Figure~\ref{fig:overview-capturing} - (4) shows pipeline results including intermediary files 
captured as the outputs in Condition 1.

@@

Since all the result files are captured in Condition 1, the tool 
is ready to label the piepeline processes based on the results in Condition 2.
However, another challenge is to label undetermined processes that read files with 
differences and write files with differences too. In this
situation, it is not possible to determine from the pipeline results 
whether the process created differences or 
was transparent.
To address this issue, we set hooks in the pipeline 
in Condition 2. Hooks are set by replacing the pipeline 
processes with a custom script. This custom script labels each process 
as \emph{transparent} or \emph{create differences} immediately after running 
the original executable. 
The replacement is done through the PATH variable, as before. Moreover, 
this custom script copies the results obtained from the same process in 
Condition 1 to the pipeline output in Condition 2 if 
it is invoked with the arguments of a process that created differences 
(Figure~\ref{fig:overview-labelling} - (3)). 
Differences are identified by comparing checksum of the output files for 
the same process in both conditions (Figure~\ref{fig:overview-labelling} - (2)).
In addition, the distance functions that are specific to file types can be added 
to our tool to measure differences more precisely. 
We developed an incremental approach that consists of the following steps:

\begin{enumerate}
  \item Start the pipeline execution in Condition 2; 
        after hooking all the processes involved in the pipeline.
  \item Label each process as \emph{transparent} or \emph{create differences} 
        right after its running.
  \item Check if process is \emph{create differences} then: 
        copy the results produced by the same process in Condition 1 to the 
        pipeline output in Condition 2. 
  \item Continue steps 2 and 3 until the end of the pipeline execution.
\end{enumerate}

This algorithm creates a \emph{labeled process tree} at the end of pipeline execution 
which highlights processes that create differences.
(Figure~\ref{fig:overview-labelling} - (4)).

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{images/iterative_modif}
  \caption{A simple example of steps of the proposed approach.}
  \label{fig:iterations}
\end{figure}

Figure~\ref{fig:iterations} illustrates our incremental labelling 
process for the example in Figure~\ref{fig:simple_script}. At every 
step, processes that created differences are shown in red and other processes 
(transparent processes) are in green. As in 
Figure~\ref{fig:simple_script}, plain black edges represent the process 
tree and dashed edges represent file dependencies: green edges 
represent files with no differences, while red edges represent files with 
differences. Temporary files are not represented because they have been 
saved as previously described.

The four steps in Figure~\ref{fig:iterations} correspond to the 
steps of the labelling algorithm. 
At step one, process \texttt{1\#bet} is executed and then labeled 
as transparent (green) as it produced files without differences.
At step 2, \texttt{2\#bet2} 
is labeled as difference creator (red) as it produced files with differences 
from files without differences. Therefore, the files produced by \texttt{2\#bet2} in  
Condition 2 are replaced with the files produced by \texttt{2\#bet2} in 
Condition 1.
At step 3, after running process \texttt{3\#fslstats}, it labeled as 
transparent as it write files without differences.
At step 4, process \texttt{4\#fslmaths} is labeled as difference creator 
as the last process of the pipeline.
As a result of these 4 steps, the final process labelling is: 
\texttt{2\#bet2} and \texttt{4\#fslmath} are difference creators (red) 
and the other processes are transparent.

Different subjects may lead to different process tree, due to
variations that may be coming from differences in data types or
cardinality. For instance, some images may have been acquired multiple
times in some subjects. Some of these differences can be neglected, for
instance when a data decompression step is present at the beginning of the
execution for some subjects only, and other ones cannot, when different
processing paths are used in different subjects.

Different process tree shows that pipeline execution followed by a different procedure. 
This implies the possible impact of different subjects on pipeline results.
We cluster subjects into different types to specify how many different 
process trees are generated. Clustering is performed based on the process trees 
which nodes are labeled by the process name. 
In this order, 
the tree edit distance~\cite{zhang1989simple} is used as a similarity measure for 
representing the distance between labeled trees. Tree edit distance  is defined  
as the minimum number of edit operations which is needed to transform one tree into 
the other. Three edit operations are considered: node label modification, 
node removal, and node insertion. Each operation has an associated cost of 1.
With this distance, we cluster subjects using agglomerative hierarchical
clustering as implemented in SciPy~\cite{oliphant2007scipy}
(Algorithm~\ref{algo:hclustering}). As a result, the distance between
any two subjects in a cluster is lower than a threshold parameter.

Furthermore, we will show that even subjects in the same cluster, 
having the same topological process tree structure, 
may have different labels as processes that create differences.

\begin{algorithm}[h!]
\caption{Hierarchical clustering algorithm from SciPy}
\label{algo:hclustering}
\begin{algorithmic}

  \State /* Input: - a list of n trees; \texttt{T = [t${_1}$, t${_2}$,...,t${_n}$]}
  \State /*\quad \quad \quad \quad - a distance threshold value; \texttt{threshold}
  \State /* Output: a list of clusters; \texttt{C = [c${_1}$,c${_2}$,...,c${_k}$]}
  \State \texttt{\# Create a cluster for each tree in input list;}
  \State \texttt{C = [ [t${_1}$], [t${_2}$], \ldots , [t${_n}$] ]}
  \State \texttt{\# Define a 2 $\times$ 2 array to store distances}
  \State \texttt{D} = \texttt{array(n, n)}
  \While{\texttt{len(C)} > \texttt{1}}  
  \State \texttt{\# Select the two nearest clusters}
  \For{\texttt{i=1} to \texttt{n}}
  \For{\texttt{j=i+1} to \texttt{len(C)}}
  \State \texttt{D[i][j]} = $\min \left\{ \texttt{edit\_dist(a, b)}: \ \texttt{a} \in \texttt{C[i]}, \texttt{b} \in \texttt{C[j]} \right\}$
    \EndFor
  \EndFor
  \State \texttt{i, j} = $\argmin \left\{ \texttt{D[i][j]}, \ \texttt{i, j} \in \llbracket 1, \texttt{len(C)}\rrbracket^2 \right\}$
  \If{\texttt{D[i][j]} > \texttt{threshold}}
  \State \Return \texttt{C}
  \Else
  \State \texttt{\# Merge C[j] into C[i]}
  \State \texttt{C[i]} = \texttt{merge}(\texttt{C[i] and C[j]})
  \State \texttt{\# Remove C[j] from C}
  \State \texttt{remove(C[j], C)}
  \EndIf
  \EndWhile
%  \State Function edit\_distance (C${_1}$, C${_2}$)
%  \State EndFunction
\end{algorithmic}
\end{algorithm}

@@

In this experiment, we analyse the numerical reproducibility of the computational pipelines 
and identify the origin of differences across the operating systems. 
Two types of differences can occur in the subjects due to the differences
in the operating systems. One is between-OS differences caused by the
operating system library updates and the other type, within-OS differences
occur as a result of the pseudo-random processes used in the pipelines.

@@

This section describes datasets and pipelines used for the analyses, 
how data processed through the pipelines, and the results.

@@

We used our tool to evaluate the reproducibility of the pipelines from the Human Connectome 
Project (\href{https://www.humanconnectome.org}{HCP}).
The HCP initiative is a significant effort to acquire and analyze 
brain connectivity data from 1200 healthy adults.
It enables the neuroscience 
research community to discover relationships between brain circuits and 
individual behaviors. This helps to understand a wide range of brain disorders.
The HCP project provides (1) database services (ConnectomeDB) for storing and 
sharing primary and processed data freely, and (2) data analysis pipelines that 
are available under an open-source license.

@@

HCP data collected from different types of imaging techniques, including 
structural imaging (sMRI), functional imaging (fMRI), and diffusion imaging (dMRI).
The structural images include T1-weighted (T1w) and T2-weighted (T2w) images which 
help in the diagnosis of the brain injury.
The images are indexed with a suffix if several scans of the same modality were acquired.
For example, some data may include two T1w images with low and high resolutions.
The functional images, including task-based fMRI and resting-state scans, 
enable the measurement of functional activations within brain areas. 
Diffusion imaging is another kind of MRI technique, which measures 
the anatomical connectivity between regions.
See~\cite{van2013wu} for a discussion of these various modalities.

Data from 1200 subjects 
is available in \href{https://db.humanconnectome.org}{the ConnectomDB repository}. 
In this paper, we randomly selected 20 unprocessed subjects 
containing the 3-tesla type of MRI scanners from the HCP data release S500.

@@

The data were processed with PreFreeSurfer and FreeSurfer. 
First, PreFreeSurfer was utilized to analyse the unprocessed data, and then its 
results were used as the input of FreeSurfer.

Each subject was processed twice on the same operating system to detect
variability coming from pseudo-random operations. 
Within-OS variability may also be coming from other factors than pseudo-randomness. 
For instance, metadata may differ due to the presence of timestamps, 
file paths, and other execution-specific information. 
To ensure that differences belong to the file data part, 
we added distance functions specific to the file types.
These functions are \emph{mri\_diff}, \emph{mris\_diff}, and \emph{lta\_diff} 
from FreeSurfer tool for comparing volumes, surface-based files, 
and lta files (transforms) respectively. 
These functions compare files based on the image dimensions, 
pixel intensities, the number of vertices, and geometry information
(see~\cite{fischl2012freesurfer} for more details). 
Moreover, we used the Dice coefficient metric to evaluate the similarity of 
the image segmentation results, which is defined by
\[DICE=\frac{2(|X|\cap|Y|)}{|X|+|Y|}\]
The coefficient between 0 to 1, 1 means that two images are totally identical, while if 
X and Y have no voxels in common, it is equal to 0.

Furthermore, we clustered subjects into different type using 
the clustering method as described before.
We set the threshold value to zero for the clustering method. 
This turns to cluster the subjects with the same topology into the same type.
Also, we used the nearest neighbor algorithm to calculate the distance between clusters.

@@

We didn't observe any within-OS difference in PreFreeSurfer, however, we
identified two FreeSurfer processes leading to within-OS differences due
to the use of pseudo-random numbers: image registration with
\emph{mri\_segreg}, and cortical surface curvature
estimations with \emph{mris\_curvature}. To address this issue, we fixed
the seed used by the pseudo-random number generator, through defining 
option ``--seed'' in the command-line of the processes.
Our analysis can focus on between-OS differences.

@@

Results show the variability between processes when 
the same tool behaves differently in different process instances. For example, 
FLIRT tool always produces files without differences in \emph{AAve} sub-pipeline but 
mostly creates different files in \emph{AAli} sub-pipeline.

Furthermore, different frequencies show the variability between subjects.
Results show that the same process can be reproducible for some subjects but not for others. 
Between subject differences can be a result of different subject types. 
For instance, the subjects contain two T1 images need more calculations 
to produce an average image between them.
The similar findings in another study~\cite{bowring2019exploring} were reported 
the effect of subject type on reproducibility.

% make figure in two columns use {figure*}
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{images/pfs_heatmap.png}
  \caption{Heatmap frequency of processes along with the sub-pipelines that introduce 
           differences among 20 subjects. 
           Each column represents the pipeline tool, and each row shows the sub-pipeline.
           Each cell may contain several process instances of a tool in a sub-pipeline.
           Annotations describe the frequency distribution of the instances as ``N [M] x Y" 
           in which N is the percentage of the process 
           that creates differences out of all M times that is occurred in the sub-pipeline. 
           Also, times Y shows the number of instances with the same distribution.
           For example, the cell of FLIRT tool in sub-pipeline AAve shows 10 process instances  
           with the frequency of 10 that introduce no differences.
           The range between red and green colors shows the highest and lowest 
           proportion of occurrences of the processes with differences.
           Also, grey cells indicate that the pipeline tool is not executed in the sub-pipeline.}
  \label{fig:pfs_freq}
\end{figure*}

Figure~\ref{fig:fnirt_result} 
shows the checkerboard images from the results of ACPC-Alignment in PreFreeSurfer. 
Each rectangle includes the replaced pixels with different intensity values between images.
The misaligned regions visually represent significant differences in the alignment 
procedure.

@@

Spot tool identified \emph{mris\_make\_surfaces} as the only process in 
the pipeline that creates differences in 10 subjects among 20 subjects.
This process generates surface files for cortical and white matter surfaces 
(see~\cite{fischl2012freesurfer} for more detail). 
Also, FreeSurfer did not introduce any differences in 10 subjects.

>>>>

Operating systems and other computational factors are known to have an
effect on the results produced by neuroimaging pipelines, presumably due to
the creation, propagation and amplification of numerical errors along data
analysis pipelines. However, the precise causes of such instabilities and
the path along which they propagate in the pipelines are unclear, while
this process could play a major role in the current reproducibility crisis.
We present a technique to identify the computational processes in a
pipeline that create numerical differences when executed in different
computational conditions. Our technique leverages system-call interception to reconstruct
processing and file graphs without pipeline instrumentation. By applying our tool to
the HCP pre-processing pipelines, 
we find that linear and non-linear registration are the cause of
most numerical instabilities in such pipelines, which confirms previous
findings.

@@

In neuroimaging, our primary application field, data analyses often consist
of hundreds of computational processes, sometimes coming from multiple
toolboxes, that are aggregated to perform a specific function. For
instance, the fMRIprep pipeline~\cite{esteban2019fmriprep} assembles software blocks
from FSL, AFNI, Freesurfer and ANTs to provide a state-of-the art
functional MRI processing tool with minimal user input. Another example are the pipelines of the Human
Connectome Project~\cite{glasser2013} that combine tools from FSL and Freesurfer to pre-process
structural, functional and diffusion data from their
open dataset. In both cases, pipelines leverage toolboxes that are
widely trusted in the community. At the same time, substantial results
variations have been observed in these toolboxes resulting from minor data
or infrastructure perturbations~\cite{Glatard2015, Gronenschild2012, Lewis2017-ll}, which suggests that further investigation of their
numerical conditioning is required. For such complex pipelines, a
lightweight solution has to be found to perform such evaluations with
limited code instrumentation.

@@

Spot identifies processes that create differences from a process graph of the pipeline execution.
This tool compares file dependencies through all the processes based on the result files in two conditions.
In this paper, different conditions refer to various operating systems.
This tool is applicable on the complex pipelines in which tracing a large number of processes is difficult.

Figure~\ref{fig:overview-tool} shows an overview of our method. For each tested
condition, Condition 1 and Condition 2, the pipeline is containerized with
Docker, and its interface is described with
Boutiques~\cite{glatard2017boutiques}. 
\reprozip is added to Condition 1 for provenance tracking. 
This captures process dependencies through system call interceptions, 
then Spot tool creates related process trees (Figure~\ref{fig:overview-tool} - (1)).
Different subjects produce different process trees. 
Hierachical clustering used to group subjects into different types (Figure~\ref{fig:overview-tool} - (2)).
The rest of the procedure is done separately on each subject type.
Our tool continues by
restoring temporary files and ``multi-write" files, i.e., files written
by more than one process (Figure~\ref{fig:overview-tool} - (3)). 
To do so, the tool creates a list of processes 
that introduce intermediary files and replaces their corresponding executables with a wrapper script. 
The tool re-executes the pipeline in a modified Docker image in Condition 1.
This copies files in the list to a read-only directory.
In the next step, the
processes are labeled from the differences found in their outputs in the
two tested conditions (Figure~\ref{fig:overview-tool} - (4)). 
The tool builds the Docker image for Condition 2 by modifying all
the processes in the pipeline. Then, the modified Docker
image is executed in Condition 2 and each process is labeled after
running by comparing the results obtained in Condition 1
and output files of the process in Condition 2. We label processes in 2
categories depending on whether they
create differences (represented in red) or not (green). 
The tool replaces output file with the same file obtained in Condition 1 
if the process creates differences.
Process labelling
is done through hooks which are defined during the Docker image modifying
procedure in Condition 2.
Spot tool is available at https://github.com/big-data-lab-team/repro-tools.

\begin{figure}
  \centering
    \includegraphics[width=1\columnwidth]{images/overview-fig}
    \caption{An overview of Spot tool in four steps.
    (\href{https://encsconcordiaca-my.sharepoint.com/:p:/g/personal/m_alari_encs_concordia_ca/EYx_aIZgYKlDuuXKPKi5ckIBo_WmJeE1Czr9rqG4gLHOOA?e=H8gack}{Source})}
    \label{fig:overview-tool}
  \end{figure}

@@

Our tool reconstructs a \emph{process tree} starting from the first process
created by the pipeline and traversing processes through \texttt{clone()}
and \texttt{fork()} dependencies. Our tool also creates a process \emph{graph} 
from this tree
by adding edges corresponding to file dependencies between processes. A
file dependency is defined between processes A and B if a file written by A
is read by B. Figure~\ref{fig:simple_script} shows an example of a process
tree and graph constructed from the example pipeline in
Listing~\ref{listing:sample-script}.

Different subjects may lead to different process trees, due to
variations that may be coming from differences in data types or
cardinality. For instance, some images may have been acquired multiple
times in some subjects. Some of these differences can be neglected, for
instance, when a data decompression step is present at the beginning of the
execution for some subjects only, and other ones cannot, when different
processing paths are used in different subjects.
Accordingly, we cluster subjects into different types.

\begin{listing}
\begin{minted}[frame=single,
  framesep=3mm,
  linenos=false,
  xleftmargin=0pt,
  tabsize=4]{bash}
#!/bin/bash

@@

# Run FSL bet, put result in ${bet_output}
bet "${input_image}" "${bet_output}" > bet_temp.out
echo "Voxels / Volume in brain mask:"
# Run FSL stats on ${bet_output}
fslstats "${bet_output}" -V
# Run FSL maths on ${bet_output}
fslmaths "${bet_output}" "${bet_output_binarized}"
# Remove temporary file
\rm bet_temp.out
\end{minted}
  \caption{Example pipeline}
  \label{listing:sample-script}
\end{listing}

@@

We cluster subjects into different types to specify how many different 
process trees are generated. 
Nodes in the process trees are labeled by the process name.
The tree edit distance~\cite{zhang1989simple} is used as a similarity measure for 
representing the distance between labeled trees. 
Tree edit distance  is defined  
as the minimum number of edit operations which is needed to transform one tree into 
the other. Three edit operations are considered: node label modification, 
node removal, and node insertion. Each operation has an associated cost of 1.
With this distance, we cluster subjects using agglomerative hierarchical
clustering as implemented in SciPy~\cite{oliphant2007scipy}.
As a result, the distance between
any two subjects in a cluster is lower than a threshold parameter.

We will show that different types of subjects certainly produce different results, 
while subjects in the same type can be either reproducible or not.

@@

We clustered subjects into different types that each one refers to a specific graph.
The processes of the graph of each subject are labeled into two categories. First,
processes that read files that do not have differences and write files that
do not have differences are labeled \emph{transparent}
(Figure~\ref{fig:green}). Second, processes that read files that do not
have differences but write files that have differences \emph{create}
differences (Figure~\ref{fig:red}). In addition, processes that read files
that have differences and write files that also have differences are called
\emph{undetermined} (Figure~\ref{fig:grey}). Processes are also labeled
undetermined when they read or write temporary files, or produce multi-write
files. To resolve undetermined processes, hooks must be added to the pipeline to 
execute these processes on files without differences, and to capture
intermediary file states.

@@

Files written by multiple processes also lead 
to undetermined labels. For a file F 
written by all processes in \textbf{P} = \{$P_{1}$, \ldots $P_{n}$\}, we 
(1) check that processes in \textbf{P} do not write concurrently to F, 
(2) establish an order on \textbf{P} based on the creation timestamp 
of the processes, (3) replace every process $P_{i}$ in \textbf{P} by 
a wrapper that first calls $P_{i}$ and then saves F to a read-only 
directory. Thus, multiple versions of F are saved and used in the 
analysis.

Furthermore, cycles may be present in the process graph in case a file 
was written by more than one process. We remove such cycles by removing 
file edges between processes A and B when A's process creation 
timestamp is posterior to B's or when A=B. Indeed, such edges cannot 
happen in practice unless A and B were running concurrently, which we 
assume is not the case.

As an example, in the sample pipeline in Listing~\ref{listing:sample-script}, process \texttt{2\#bet2} writes a temporary file.
Also, file \emph{T1w\_brain\_bin.nii.gz} is first written by \texttt{2\#bet2} and then overwritten by process \texttt{3\#fslmaths}.
The complete list of the processes that write intermediary files is captured in 
Listing~\ref{listing:json_process}.
Using this list, we modify processes in Condition 1 and then re-execute pipeline 
to save them in a backup folder.

@@

When all the result files are captured in Condition 1, the tool 
is ready to label the pipeline processes in Condition 2.
However, another challenge is to label undetermined processes that read files with 
differences and write files with differences too. 
In this case, it is not possible to determine from the pipeline results 
whether the process creates differences or 
is transparent.
To address this issue, we set hooks in the pipeline 
in Condition 2. Hooks are set by replacing the pipeline 
processes with a wrapper script.
The replacement is done by modifying the PATH environment variable, as explained before. 
The wrapper script labels each process 
as \emph{transparent} or \emph{create differences} after running 
the original executable. 
Then, it copies the results obtained from the same process in 
Condition 1 to the pipeline output in Condition 2 if 
it is invoked with the arguments of a process that created differences.
Differences are identified by comparing the checksum of the output files for 
the same process in both conditions.
Besides, to measure the magnitude of the differences, we can add distance functions specific to file types.
We developed an incremental approach that consists of the following steps:

\begin{enumerate}
  \item Start the pipeline execution in Condition 2; 
        after hooking all the processes involved in the pipeline.
  \item Label each process as \emph{transparent} or \emph{create differences} 
        after running.
  \item Check if process is \emph{create differences} then: 
        copy the results produced by the same process in Condition 1 to the 
        pipeline output in Condition 2. 
  \item Continue steps 2 and 3 until the end of the pipeline execution.
\end{enumerate}

This algorithm creates a \emph{labeled process tree} at the end of pipeline execution 
which highlights processes that create differences.

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{images/iterative_modif}
  \caption{Illustration of the labelling approach.}
  \label{fig:iterations}
\end{figure}

Figure~\ref{fig:iterations} illustrates our incremental labelling 
process for the example in Figure~\ref{fig:simple_script}. At every 
step, processes that created differences are shown in red and other processes 
(transparent processes) are in green. Temporary files are not represented 
because they have been saved as previously described.

The four steps in Figure~\ref{fig:iterations} correspond to the 
steps of the labelling algorithm. 
At step one, process \texttt{1\#bet} is executed and then labeled 
as transparent (green) as it produced files without differences.
At step 2, \texttt{2\#bet2} 
is labeled as difference creator (red) as it produced files with differences 
from files without differences. Therefore, the files produced by \texttt{2\#bet2} in  
Condition 2 are replaced with the files produced by \texttt{2\#bet2} in 
Condition 1.
At step 3, after running process \texttt{3\#fslstats}, it labeled as 
transparent as it writes files without differences.
At step 4, process \texttt{4\#fslmaths} is labeled as difference creator 
as the last process of the pipeline.
As a result of these 4 steps, the final process labelling is: 
\texttt{2\#bet2} and \texttt{4\#fslmath} are difference creators (red) 
and the other processes are transparent.

@@

In this experiment, we analyse the numerical reproducibility of computational pipelines 
and identify the origin of differences across the operating systems. 
Two types of differences can occur in the subjects due to the differences
in the operating systems. One is between-OS differences caused by the
operating system library updates and the other type, within-OS differences
occur as a result of the pseudo-random processes used in the pipelines.

@@

This section describes datasets and pipelines used for the analyses, and the way of data processing.

@@

We used our tool to evaluate the reproducibility of the pipelines from the Human Connectome 
Project (\href{https://www.humanconnectome.org}{HCP}).
The HCP initiative is an effort to acquire and analyse 
brain connectivity data from 1200 healthy adults.
It enables the neuroscience 
research community to discover relationships between brain circuits and 
individual behaviors. This helps to understand a wide range of brain disorders.
The HCP project provides database services (ConnectomeDB) for storing and 
sharing primary and processed data freely, and data analysis pipelines that 
are available under an open-source license.

@@

HCP data collected from different types of imaging techniques, including 
structural imaging (sMRI), functional imaging (fMRI), and diffusion imaging (dMRI).
The structural images include T1-weighted (T1w) and T2-weighted (T2w) images, which 
help in the diagnosis of brain injury.
The images are indexed with a suffix if several scans of the same modality were acquired.
For example, some data may include two T1w images with low and high resolutions.
The functional images, including task-based fMRI and resting-state scans, 
enable the measurement of functional activations within brain areas. 
Diffusion imaging is another kind of MRI technique, which measures 
the anatomical connectivity between regions.
See~\cite{van2013wu} for a discussion of these various modalities.

In this paper, we randomly selected 20 unprocessed subjects 
containing 3-tesla type of MRI scanners from the HCP data release S500 
available in \href{https://db.humanconnectome.org}{the ConnectomDB repository}.

@@

The data were processed with PreFreeSurfer and FreeSurfer. 
First, PreFreeSurfer was utilized to analyse the unprocessed data, and then its 
results were used as the input of FreeSurfer.
Each subject was processed twice on the same operating system to detect
variability coming from pseudo-random operations. 
Within-OS variability may also be coming from other factors than pseudo-randomness. 
For instance, the same files may have different metadata such as timestamps, 
file paths, and other execution-specific information. 
To ensure that the differences are not related to the metadata, 
we used specific functions from FreeSurfer tool including 
\emph{mri\_diff}, \emph{mris\_diff}, and \emph{lta\_diff}.
These functions compare files based on the image dimensions, 
voxel intensities, the number of vertices, and geometry information
(see~\cite{fischl2012freesurfer} for more details). 
In addition, we used the Dice coefficient metric to evaluate the similarity of 
the image segmentation results, which is defined by
\[DICE=\frac{2(|X|\cap|Y|)}{|X|+|Y|}\]
The Dice coefficient ranges from 0, indicating X and Y have no voxels in common, to 1, if 
two images are identical.

Finally, to cluster the subjects, the threshold value is set to zero in the clustering method. 
Also, we used the nearest neighbor algorithm to calculate the distance between clusters.

@@

We did not observe any within-OS difference in PreFreeSurfer, however, we
identified two FreeSurfer processes leading to within-OS differences due
to the use of pseudo-random numbers: image registration with
\emph{mri\_segreg}, and cortical surface curvature
estimations with \emph{mris\_curvature}. To address this issue, we fixed
the seed used by the pseudo-random number generator, through defining 
option ``--seed'' in the command-line of the processes.
Our analysis can focus on between-OS differences.

@@

The results show two types of differences, between-process and between-subject.
Between-process differences occur when the same tool behaves differently in different process instances.
For example, 
FLIRT tool always produces files without differences in \emph{AAve} sub-pipeline but 
mostly creates different files in \emph{AAli} sub-pipeline.
Also, different frequencies show between-subjects variability.
Accordingly, the same process can be reproducible for some subjects but not for others. 
The similar findings in another study~\cite{bowring2019exploring} were reported 
the effect of between-subject differences on reproducibility.
% Between subject differences can be a result of different subject types. 
% For instance, the subjects contain two T1 images need more calculations 
% to produce an average image between them.

% make figure in two columns use {figure*}
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{images/pfs_heatmap.png}
  \caption{Frequency heatmap of processes in the tools that introduce differences among 20 subjects. 
           Each column represents the pipeline tool, and each row shows the sub-pipeline.
           Each cell may contain several instances of the process of a tool in a sub-pipeline.
           Annotations describe the frequency distribution of the instances as ``N [M] x Y" 
           in which N is the proportion of the process 
           that creates differences out of all M times that is occurred. 
           Also, times Y shows the number of processes with the same distribution.
           For example, the cell of FLIRT tool in the sub-pipeline AAve shows 10 instances of the process 
           with the frequency of 11 that introduce no differences.
           The range between red and green colors shows the highest and lowest 
           proportion of occurrences of the processes with differences.
           Also, grey cells indicate that the tool is not executed in the sub-pipeline.}
  \label{fig:pfs_freq}
\end{figure*}

Figure~\ref{fig:fnirt_result} 
shows checkerboard image from the results of ACPC-Alignment in PreFreeSurfer. 
Each rectangle includes the replaced voxels with different intensity values between images.
The misaligned regions visually represent significant differences in the alignment 
procedure.

@@

We processed 20 subjects using FreeSurfer and 
Spot tool identified \emph{mris\_make\_surfaces} as the only process in 
the pipeline that creates differences in 10 subjects.
This process generates surface files for cortical and white matter surfaces 
(see~\cite{fischl2012freesurfer} for more detail). 
Also, FreeSurfer did not introduce any differences in 10 subjects.
